Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-09-03 manuscript/main.tex
better yet, an appendix

\section{Introduction}\label{sec:intro}

@@

\section{Model}\label{sec:model}

The setting for our modeling is Kingman's coalescent \citep{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
In the Methods we detail our retracing of the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
We make use of the results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} to facilitate computation, and build on the notation of \cite{Rosen2018-bb} for a finite-dimensional approximation of our demographic and mutation intensity histories.
Complete derivation of formulae are found in the Methods.

Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Methods section \ref{sec:methods:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
\begin{equation}
  \label{eqn:transform}
\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu,
\end{equation}
where $\mathcal{L}_{n,\eta}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
It is dependent on $n$ and $\eta$, which we take to be fixed.

@@

For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\mu(t)$ as piecewise constant on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
We take the epoch boundaries as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $\boldsymbol z = (z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
In Methods section \ref{sec:methods:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
\begin{equation}
\label{eqn:transform_discrete}
\boldsymbol \xi = L_{n, \boldsymbol y} \boldsymbol z,
\end{equation}
where the $(n-1)\times m$ matrix $L_{n, \boldsymbol y}$ is fixed given a fixed demographic history $\boldsymbol y$.

@@

\section{Results}\label{sec:results}

@@

\subsection{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}

@@

From \eqref{eqn:pi} and \eqref{eqn:r} in the Methods, the CDF of $T_2$ is
\begin{align}
F_2(t) &\equiv 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
\end{align}

\section{Discussion}\label{sec:discussion}

@@

\section{Methods}\label{sec:methods}

\subsection{The expected SFS given demographic and mutation intensity histories}\label{sec:methods:xi}

Suppose $n$ haplotypes are sampled in the present, and let rvs $(T = T_2,\dots,T_n)$ denote the coalescent times measured retrospectively from the present (i.e. $T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
This is tantamount to the infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
%WD Kam suggests replacing below with $\mu^* \ge \mu(t) \ge 0$, but I don't get why
Further suppose that mutation intensity is not constant, but a specified function of time $0\le \mu(t)<\infty$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
A given line in the coalescent tree then acquires mutations on a genomic subinterval $(x,x+\delta x)$ at rate $\mu(t)\delta x$.
The following argument can be generalized to allow the labelling distribution to be nonuniform over the unit interval, but the same results follow.

Let rv $\mathcal{E}_{\delta x, b}$ denote the event that a mutation present in $b\in\{1, 2, \dots, n-1\}$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (the probability of this event is independent of $x$, given the uniform labeling distribution).
Let $I_k$ denote the $k$th intercoalescent time interval, i.e.\ $I_n = (0, T_n),\ I_{n-1} = (T_n, T_{n-1}),\ \dots,\ I_2 = (T_3, T_2)$.
Let rv $\mathcal{E}_{\delta x, b, k}$ denote the event that the mutation $\mathcal{E}_{\delta x, b}$ occurred during $I_k$.
For small $\delta x$ and finite $\mu(t)$ we have
\begin{align*}
\mathbbm{P}(\mathcal{E}_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(\mathcal{E}_{\delta x, b, k}\mid T)\\
&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{t\in I_k}\mu(t)dt + O((\delta x)^2)\right),
\end{align*}
where
\begin{equation}
\label{eqn:p}
p_{n,k}(b) = \frac{\binom{n-b-1}{k-2}}{\binom{n-1}{k-1}}
\end{equation}
is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parentheses is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
Marginalizing $T$ gives
\begin{align*}
\mathbbm{P}(\mathcal{E}_{\delta x, b}) &= \delta x\sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right] + O((\delta x)^2).
\end{align*}
For small $\delta x$, each such interval contains zero or one mutations, so the expected number of mutations subtending $b$ haplotypes (i.e.\ the $b$th component of the SFS) is
\[
\xi_b = \int_0^1 \mathbbm{P}(\mathcal{E}_{dx, b})dx = \sum_{k=2}^n k p_{n,k}(b)\mathbbm{E}_T\left[\int_{t\in I_k}\mu(t)dt\right]\nonumber
\]
Using the explicit bounds of the intercoalescent intervals, $I_{k} = (T_{k+1}, T_k)$, gives
\begin{align}
\label{eqn:xi}
\xi_b &= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=2}^{n-1} k p_{n,k}(b) \mathbbm{E}_{T_{k+1}}\left[\int_0^{T_{k+1}}\mu(t)dt\right]\nonumber\\
&= \sum_{k=2}^n k p_{n,k}(b) \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \sum_{k=3}^{n} (k-1) p_{n,k-1}(b) \mathbbm{E}_{T_{k}}\left[\int_0^{T_k}\mu(t)dt\right]\nonumber\\
&= \sum_{k=2}^n B_{b,k} \mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right],
\end{align}
where
\begin{equation}
\label{eqn:B}
B_{b,k}\equiv
\begin{cases}
k p_{n,k}(b),& k=2\\
k p_{n,k}(b) - (k-1) p_{n,k-1}(b),& k > 2
\end{cases}
\end{equation}

\cite{Polanski2003-kg} (eqns 5-8) give the marginal density for the coalescent time $T_k$ as
\begin{equation}
\label{eqn:pi}
\pi_k(t_k) = \sum_{j=k}^n A_{k,j} q_j(t_k)
\end{equation}
where
\begin{align*}
A_{k,j} &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
A_{n,n} &\equiv 1,\\
q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
\end{align*}
and $\eta(t)$ is the haploid effective population size history.
Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present, with inhomogeneous Poisson intensity function $\binom{j}{2}/\eta(t)$.

The expectations in \eqref{eqn:xi} can be expressed using \eqref{eqn:pi} as
\begin{align}
\label{eqn:exp}
\mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] &= \int_0^\infty\pi_k(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
&= \sum_{j=k}^n A_{k,j}\int_0^\infty q_j(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
&= \sum_{j=k}^n A_{k,j}\int_0^\infty q_j(t_k)\int_0^\infty 1_{[0<t<t_k]}(t)\mu(t)dt dt_k\nonumber\\
&= \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt
\end{align}
where in the last line we've exchanged integration order and defined the inhomogeneous Poisson survival function
\begin{equation}
\label{eqn:r}
r_j(t) \equiv \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]
\end{equation}
corresponding to density $q_j(t)$ (that this is the appropriate survival function can be seen by considering the limit of a sequence of Bernoulli failures).

Using \eqref{eqn:exp} in \eqref{eqn:xi} gives
\begin{align}
\label{eqn:xi2}
\xi_b &= \sum_{k=2}^n B_{b,k} \sum_{j=k}^n A_{k,j}\int_0^\infty r_j(t)\mu(t)dt\nonumber\\
&= \sum_{j=2}^n \left(\sum_{k=2}^j B_{b,k} A_{k,j}\right) \int_0^\infty r_j(t)\mu(t)dt,
\end{align}
where we've exchanged summation order in the last line.

We then have a linear expression for the expected SFS as a function of the mutation intensity history $\mu(t)$:
\begin{equation}
\label{eqn:xivec}
\boldsymbol\xi = C \boldsymbol d,
\end{equation}
where the $(n-1)\times(n-1)$ matrix
\[
C_{b,j} \equiv \sum_{k=2}^j B_{b,k} A_{k,j}
\]
is constant wrt $\mu$ \emph{and} $\eta$, and
\begin{equation}
\label{eqn:d}
d_j \equiv \int_0^\infty r_j(t)\mu(t)dt = \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt
\end{equation}
is a linear functional of $\mu$ and a nonlinear functional of $\eta$.
We recover \ref{eqn:transform} by defining the operator $\mathcal{L}_{n,\eta}$ such that $\mathcal{L}_{n,\eta}\mu \equiv C \int_0^\infty \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]\mu(t)dt$.

\subsection{Computing the elements of $C$}\label{sec:methods:C}

%WD Kam suggests just doing C = B A, but computing the elements of B and A would seem computationally infeasible
We next develop an efficient recursive procedure for computing the $(n-1)\times(n-1)$ matrix $C$.
Using \eqref{eqn:B}
\begin{align*}
C_{b,j} &= \sum_{k=2}^j k p_{n,k}(b) A_{k,j} - \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}\\
&= W_{b,j}^{(1)} - W_{b,j}^{(2)},
\end{align*}
where
\begin{align}
\label{eqn:W1}
W_{b,j}^{(1)} &\equiv \sum_{k=2}^j k p_{n,k}(b) A_{k,j}\\
\label{eqn:W2}
W_{b,j}^{(2)} &\equiv \sum_{k=3}^j (k-1) p_{n,k-1}(b) A_{k,j}.
\end{align}
\cite{Polanski2003-ll} (eqn 11) show that $A$ can be expressed as
\[
A_{k,j} = \frac{n! (n-1)!}{(j+n-1)! (n-j)!} \frac{(2 j-1)}{j (j-1)} \frac{(j+k-2)!}{ (k-1)! (k-2)! (j-k)! }(-1)^{j-k},
\]
so, given the form of $p_{n,k}(b)$ in \ref{eqn:p} it's clear that \eqref{eqn:W1} and \eqref{eqn:W2} are definite sums over geometric terms.
Zeilberger's algorithm, which finds polynomial recurrences for definite sums of hypergeometric terms \citep{petkovvsek1996b, paule1995mathematica}, can thus be used to yield the following procedurally generated second-order recursions in $j$:
%WD need to wrap this eqn, as it runs will beyond the margin. Also make an appendix at some point for these
\begin{align*}
&W_{b,2}^{(1)} = \frac{6}{(n+1)}\\
&W_{b,3}^{(1)} = \frac{10(5n-6b-4)}{(n+2)(n+1)}\\
&W_{b,j+2}^{(1)} = -\frac{(2 j+3) \left(-(2 j-1) W_{b,j+1}^{(1)}  \left(2 j (j+1) \left(b^2 \left(j^2+j-2\right)-6 b-j (j+1)-2\right)-j (j+1) n \left(3 b \left(j^2+j+2\right)+j^2+j-2\right)+\left(j (j+1) \left(j^2+j+6\right)+4\right) n^2+4 n\right)-(j-1) (j+1)^2 (j-n) W_{b,j}^{(1)}  (4 (n+1)-j (j+2) (b-n-1))\right)}{j^2 (j+2) (2 j-1) (j+n+1) \left(-b j^2+b+\left(j^2+3\right) (n+1)\right)}
\end{align*}
and
\begin{align*}
&W_{b,2}^{(2)} = 0\\
&W_{b,3}^{(2)} = \frac{20 (n-2)}{(n+1)(n+2)}\\
&W_{b,j+2}^{(2)} = \frac{(2 j+3) (j-n+1)}{j} \left(\frac{(j+1)}{(2 j-1) (j+n)}W_{b,j}^{(2)}-\frac{(j (j+1) (2 b-n+1)-2 (n+1))}{(j-1) (j+2) (j-n) (j+n+1)}W_{b,j+1}^{(2)}\right)
\end{align*}
This completes the framing of mutation intensity inference as a linear inverse problem.

\subsection{Finite-dimensional parameterization of $\eta(t)$ and $\mu(t)$}\label{sec:methods:pcsws}

%WD Kam describes this discretization as ad hoc, but I'm pretty sure this is your textbook Riemann sum with rectangular partitions. We could do fancier quadrature, but then we'll be less able to lean on previous work (i.e. Rosen et al.)

Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \eqref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)d\tau,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:methods:xi}).
We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
%WD Kam hates the latin loooooooool. It really does follow by dropping in the y vector and proceeding as in their proof
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}
\begin{bmatrix}
y_1 &     &      &             &       \\
    & y_2 &      &             &       \\
     &     & y_3 &             &       \\
     &      &    & \ddots      &       \\
     &      &        &  & y_m
\end{bmatrix}.
\end{equation}
Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.

Combining \ref{eqn:d3} with \ref{eqn:xivec} gives the discretized inverse problem
\begin{equation}
\boldsymbol\xi = C M(\boldsymbol y) \boldsymbol z = L_{n, \boldsymbol y} \boldsymbol z,
\end{equation}
where $L\equiv C M(\boldsymbol y)$.

\subsection{Implementation and pipeline}\label{sec:methods:tool}

@@

\section{Acknowledgements}\label{sec:ack}

@@

>>>>

\section*{Introduction}\label{sec:intro}

@@

\section*{Model}\label{sec:model}

The setting for our modeling is Kingman's coalescent \citep{Kingman1982-ge, Kingman1982-tf, Kingman1982-ys, Kingman2000-jr}, with all the usual niceties: neutrality, infinite sites, linkage equilibrium, and panmixia.
In the Methods we detail our retracing of the derivation by \cite{Griffiths1998-qf} of the frequency distribution of a derived allele conditioned on the demographic history, while generalizing to a time inhomogeneous mutation process.
We make use of the results of \cite{Polanski2003-kg} and \cite{Polanski2003-ll} to facilitate computation, and build on the notation of \cite{Rosen2018-bb} for a finite-dimensional approximation of our demographic and mutation intensity histories.
Complete derivation of formulae are found in the Appendix.

Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Appendix \ref{sec:appendix:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
\begin{equation}
  \label{eqn:transform}
\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu,
\end{equation}
where $\mathcal{L}_{n,\eta}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
It is dependent on $n$ and $\eta$, which we take to be fixed.

@@

For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\mu(t)$ as piecewise constant on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
We take the epoch boundaries as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $\boldsymbol z = (z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
In Appendix \ref{sec:appendix:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
\begin{equation}
\label{eqn:transform_discrete}
\boldsymbol \xi = L_{n, \boldsymbol y} \boldsymbol z,
\end{equation}
where the $(n-1)\times m$ matrix $L_{n, \boldsymbol y}$ is fixed given a fixed demographic history $\boldsymbol y$.

@@

\section*{Results}\label{sec:results}

@@

\subsection*{Tempora incognita: observability toward the coalescent horizon}\label{sec:model:loss}

@@

From \eqref{eqn:pi} and \eqref{eqn:r} in the Methods, the CDF of $T_2$ is
\begin{align}
F_2(t) &= 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
\end{align}

\section*{Discussion}\label{sec:discussion}

@@

\section*{Methods}\label{sec:methods}

\subsection*{Implementation and pipeline}\label{sec:methods:tool}

@@

\section*{Acknowledgements}\label{sec:ack}

@@

\appendix
\input{appendix}
